

## Data example 1: Prognostic factors for body fat  
\scriptsize (From Theo Gasser \& Burkhardt Seifert \emph{Grundbegriffe der Biostatistik})

Body fat is an important indicator for overweight, but difficult to measure. 

__Question:__  Which factors allow for precise estimation (prediction) of body fat?  

Study with 241 male participants. Measured variable were, among others, body fat (\%), age, weight, body size, BMI, neck thickness and abdominal girth.  

```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(dplyr)
library(here)
d.bodyfat <- read.table(here("3_datasets/bodyfat.clean.txt"),header=T)
d.bodyfat <- d.bodyfat[,c("bodyfat","age","gewicht","hoehe","bmi","neck","abdomen")]
```

```{r echo = TRUE}
glimpse(d.bodyfat)
```

## Scatterplots

\setkeys{Gin}{width=0.85\textwidth}

```{r, fig.width=6, fig.height=6, out.width="6cm", fig.align = 'center'}
pairs(d.bodyfat)
```

## 


```{r echo = FALSE, fig=TRUE, out.width = "6cm", out.height='6cm', fig.align='center', include = TRUE}
par(mfrow=c(2,2),mar=c(4,4,1,1))
plot(bodyfat ~ bmi ,data=d.bodyfat)
plot(bodyfat ~ gewicht ,data=d.bodyfat,ylab="")
plot(bodyfat ~ neck ,data=d.bodyfat)
plot(bodyfat ~ abdomen ,data=d.bodyfat,ylab="")
```

We are looking for a _model_ that \alert{predicts} body fat as precisely as possible from variables that are easy to measure. 

## Data example 2: Mercury (Hg) in Valais (Switzerland)

__Question:__ Association between Hg concentrations in the soil and in urine of the people living in the respective properties. We use a slightly modified data set here.

```{r echo = FALSE}
d.hg <- read.table(here("3_datasets/hg_urine_fuzzed.csv"),header=T, sep=",")
#d.hg <- d.hg[,-c(11)]
#names(d.hg) <- c("Hg_urin", "Hg_soil", "veg_garden", "migration", "smoking","amalgam", "age", "fish", "last_time_fish", "mother")
d.hg$mother <- as.factor(d.hg$mother)
set.seed(234)
d.hg$Hg_soil <- d.hg$Hg_soil[sample(1:nrow(d.hg))]
```
\  

\tiny
```{r echo = TRUE}
glimpse(d.hg)
```

## 

A first visual inspection is not very informative. There is not much that is visible by eye:
\  

```{r,echo=FALSE, out.width='6cm', fig.align='center', message=FALSE, warning=FALSE}
library(ggplot2)
ggplot(d.hg,aes(x=Hg_soil,y=Hg_urin)) + geom_point() + 
  theme_bw() +
  xlab("Hg soil") +
  ylab("Hg urine")
```

## 

Which other factors might be responsible for high Hg concentrations in urine?
\  

```{r echo=FALSE, out.width='9cm', fig.align='center', message=FALSE, warning=FALSE}
require(cowplot)
theme_set(theme_cowplot(font_size=10)) # reduce default font size

plot.1 <- ggplot(d.hg,aes(x=amalgam,y=Hg_urin)) + geom_point(size=0.5) + 
  theme_bw() +
  xlab("Amalgam") +
  ylab("Hg urine")

plot.2 <- ggplot(d.hg,aes(x=age,y=Hg_urin)) + geom_point(size=0.5) + 
  theme_bw() +
  xlab("Age") +
  ylab("Hg urine")

plot.3 <- ggplot(d.hg,aes(x=fish,y=Hg_urin)) + geom_point(size=0.5) + 
  theme_bw() +
  xlab("Fish consumption") +
  ylab("Hg urine")

plot_grid(plot.1, plot.2, plot.3, ncol=3)
```

From these plots it is hard to tell which factors exactly influence the Hg pollution in humans.

## 

It is always useful to look at the distribution of the variables in the model. Let us plot the histogram of Hg concentrations:  
\  

```{r out.width='6cm', echo=FALSE, fig.align='center', message=FALSE, warning=FALSE}
plot.1 <- ggplot(d.hg,aes(x=Hg_soil)) + geom_histogram() + xlab("Hg soil")
plot.2 <- ggplot(d.hg,aes(x=Hg_urin)) + geom_histogram() + xlab("Hg urine")
plot_grid(plot.1, plot.2,ncol=2)
```

All Hg values seem to "stick" at 0.

## 

The scatterplot does also look much more reasonable with log-transformed values:
\  

```{r out.width='7cm', echo=FALSE, fig.align='center', message=FALSE, warning=FALSE}
plot(log(Hg_urin) ~ log(Hg_soil), data=d.hg, xlab="log(Hg soil)", 
     ylab = "log(Hg Urin)",pch=21,bg=as.numeric(mother)+2,xlim=c(-4.5,4.5))
legend("topright",legend=c("Children","Mothers"),col=c(3,4),pch=21,pt.bg=c(3,4))
```

Remember: The idea to log-transform the variables was mainly obvious thanks to \alert{visual inspection}! 

## Data example 3: Diet and blood glucose level

<!--{\scriptsize\citep[p. 190]{elpelt.hartung1987}}-->

24 persons were split into 4 groups. Each group followed another diet (DIAET). The blood glucose concentrations were measured at the beginning and at the end (after 2 weeks). The difference of these values was stored (BLUTZUCK).  

\colorbox{lightgray}{\begin{minipage}{14cm}
{\bf Question:} Are there differences among the groups with respect to changes in blood glucose concentrations?
\end{minipage}}

Let's look at the raw data (points and boxplots):

```{r echo = FALSE}
path <- "../../../3_datasets/"
d.blz <- read.table(here("3_datasets/blutzucker.dat"),header=T)
```

```{r out.height='4cm', out.width='4cm',echo=FALSE, fig.align='center'}
ggplot(d.blz,aes(x=factor(DIAET),y=BLUTZUCK)) + geom_boxplot() + geom_point(size=3,colour='grey') + theme_bw() + xlab("Diet") + ylab("Blood glucose")
```

## Data example 4: Blood-screening

<!--{\scriptsize\citep[From ][Chapter 7.1]{hothorn.everitt2014}}-->

Is a high ESR (erythrocyte sedimentation rate) an indicator for certain diseases (rheumatic disease, chronic inflammations)?  

__Specifically:__ Is there an association between the concentrations of the plasma proteins Fibrinogen and Globulin and ESR level ESR$<20mm/hr$?

The `plasma` data come with the package `HSAUR3` <!--\citep{hothorn.everitt2014}.-->

```{r echo = FALSE, message=FALSE, warning=FALSE}
library("HSAUR3")
data("plasma",package="HSAUR3")
plasma$status <- factor(ifelse(plasma$ESR=="ESR < 20","healthy","sick"))
plasma[c(1,5,9,10,15,29),]
```

## 

The distinction ESR$<20mm/hr$ (healthy) vs.\ ESR$\geq 20mm/hr$ (sick) leads to a \alert{binary} response variable.  

\begin{center}$\rightarrow$ \emph{conditional density plot}\end{center}

```{r out.width='7cm', out.height='4cm', fig.align='center'}
par(mfrow=c(1,2))
cdplot(status ~ fibrinogen,plasma)
cdplot(status ~ globulin,plasma)
```




## The scopes of statistical data analysis

a. \alert{Prediction (extrapolation), interpolation}. Example body fat: use substitute measurements to predict body fat of a person.

b. \alert{Explanation;  determination of important variables}. Example physical activity of children: The study aims to find factors that (positively or negatively) influence the movement behavior of children.

c. \alert{Estimation of parameters and quantify the uncertainty.} Example: Effect size of a novel drug.

d. Optimization.

e. Calibration.

In this course we are concerned with a-c.


## Coplots

Ideal to graphically display dependencies when more than two variables are involved. Very useful for categorical variables. Example: Mercury in Valais.

```{r out.width='6cm', out.height='5cm'}
#coplot(log(Hg_urin) ~  age | mother * migration,
# d.hg, panel = panel.smooth)
```
