---
title: "Course Bio144: Data Analysis in Biology"
subtitle: "Lecture 1: Introduction and Outlook"
author: "Owen Petchey"
institute: "University of Zurich"
date: "`r format(Sys.time(), '%d %B, %Y')`"
output:
  beamer_presentation:
    includes:
      in_header: ../../beamer_stuff/preamble.tex
bibliography: refs.bib
classoption: "aspectratio=169"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
options(xtable.comment = FALSE)
```

## Literature

Compulsory literature (books available as ebooks from uzh):

1. _Lineare Regression_ by W. Stahel (pdf on course webpage)
2. _Getting Started with R, An introduction for biologists_ (__Second Edition__) Beckerman, Childs \& Petchey, Oxford University Press (DO NOT USE THE FIRST EDITION!).
3. _The New Statistics With R_ by A. Hector, Oxford University Press;  

$\rightarrow$ See ``Course texts/material'' on course website.

\begin{center}
\includegraphics[width=2.7cm]{pictures/petchey_buch.jpeg} \qquad \qquad
\includegraphics[width=2.7cm]{pictures/hector.jpeg}
\end{center}

## Schedule (12 lecture units + 2 self-study weeks) {.columns-2}

\begin{multicols}{2}
{\bf Unit 1} Introduction and outlook \\[1mm]
{\bf Unit 2} No lecture \\[1mm]
{\bf Unit 3} Simple linear regression\\[1mm]
{\bf Unit 4} Residual analysis, model validation\\[1mm]
{\bf Unit 5} Multiple linear regression \\[1mm]
{\bf Unit 6} ANOVA  \\[1mm]
{\bf Unit 7} ANCOVA; Matrix Algebra \\[5mm]
{\bf Unit 8} Model selection  \\[1mm]
{\bf Unit 9} Interpretation of results, causality \\[1mm]
{\bf Unit 10} Count data (Poisson regression) \\[1mm]
{\bf Unit 11} Binary Data (logistic regression)\\[1mm]
{\bf Unit 12} Measurement error, random effects, selected topics \\[1mm]
\end{multicols}

## Overarching goals of the course



* Provide a \textcolor{blue}{solid foundation} for answering biological questions with quantitative data.  
* Help students to understand the \textcolor{blue}{language of a statistician.}  
* Ability to understand and interpret results \textcolor{blue}{in research articles.}  
* Give the students a \textcolor{blue}{challenging, engaging, and enjoyable} learning experience.  

\  

My belief: A solid foundation in statistics makes you independent!  

## Why is statistical data analysis so relevant for the biological and medical sciences?  

Awareness that, without a profound knowledge in statistical data analysis, it will be hard to analyze your data from Bachelor, Master or PhD theses.... 

* \textcolor{blue}{Medicine:} What is the effect of a drug? Which factors cause cancer?
* __Ecology:__ What is a suitable habitat for a certain animal? Which resources does it need or prefer?
* \textcolor{red}{Evoloutionary biology:} Do highly inbred animals have decreased chances to survive or reproduce?  

## Learning by doing??  

$\rightarrow$ __Not advisable__ in statistics. Experience is essential, there are many pitfalls.

A good foundation in statistics \textcolor{blue}{makes you more independent} from consultants or the goodwill of colleagues. Without such a knowledge, you will always need help from others.

Data analysis/statistics is itself an exciting part of research! 
 
Data analysis is at the \textcolor{blue}{interface between mathematics and biology/medicine} (and many other applied research fields).


## Own examples

\alert{Otter (lutra lutra)}

\colorbox{lightgray}{\begin{minipage}{14cm}
\textit{Research questions}: What is the preferred habitat by otters? How do otters adapt to human altered landscapes?
\end{minipage}}

_Method:_ Study in Austria, 9 Otter were radio-tracked and monitored during 2-3 years.  

\includegraphics[width=10cm]{pictures/otters.jpeg}
\scriptsize http://www.prolutra.ch/

## 

\alert{Inbreeding in Alpine ibex}

\colorbox{lightgray}{\begin{minipage}{14cm}
\emph{Research question:} Does inbreeding in Alpine ibex populations have a negative effect on long-term population growth? Inbreeding depression!
\end{minipage}}  

\begin{multicols}{2}
\emph{Methods:} Genetic information from blood samples allow to quantify the level of inbreeding in each ibex population. In addition, long-term monitoring of population sizes and harvest rates.  

\includegraphics[width=4cm]{pictures/steinbock.jpg}

\includegraphics[width=4cm]{pictures/ibex_graph.png}
\end{multicols} 

## \alert{Mercury (Hg) in the soil}

\begin{center}
\includegraphics[width=8cm]{pictures/wallis.png}
\end{center}

\colorbox{lightgray}{\begin{minipage}{14cm}
\emph{Research question:} Is the Hg level in the environment (soil) of people's homes associated to the Hg levels in their bodies (urin, hair)?
\end{minipage}}

_Method:_ Measurements of Hg concentrations on people's properties, as well as measurements and survey of children and their mothers living in these properties.

Highly delicate, emotionally charged, political question!
\href{http://www.srf.ch/news/regional/bern-freiburg-wallis/quecksilber-im-walliser-boden-schadete-der-gesundheit-nicht}
{\beamergotobutton{Schweiz Aktuell, 20. Juni 2016}}

## 

\alert{Physical activity in children (Splashy study)}

\begin{center}
\includegraphics[width=6cm]{pictures/kids.jpg}  

{\scriptsize splashy.ch}
\end{center}

\colorbox{lightgray}{\begin{minipage}{14cm}
\emph{Research question:} Which factors influence physical activity patterns in children aged 2-6 years?
\end{minipage}}

_Method:_ The children had to wear accelerometers for several days. In addition, their parents had to fill in a detailed questionnaire.  

Observed variables were, e.g., media consumption, behavior of the parents, age, weight, social structure,...  

## Statistics in the news (April 2016)

\begin{center}
\includegraphics[width=9cm]{pictures/NZZ1.jpeg}
\end{center}

## Producing nonsense with statistics... 

... is too easy ...  

\  

A profound knowledge of data analysis and statistics protects you from producing nonsense -- and helps to detect it. See for example:  

\ 

\textcolor{blue}{\href{https://medium.com/wintoncentre/the-risks-of-alcohol-again-2ae8cb006a4a}{The risks of alcohol (by David Spiegelhalter, 23.\ August 2018)}}   
\ 

\textcolor{blue}{\href{http://callingbullshit.org/syllabus.html}{``Calling bullshit'' course (University of Washington)}}

## Data example 1: Prognostic factors for body fat  
\scriptsize (From Theo Gasser \& Burkhardt Seifert \emph{Grundbegriffe der Biostatistik})

Body fat is an important indicator for overweight, but difficult to measure. 

__Question:__  Which factors allow for precise estimation (prediction) of body fat?  

Study with 241 male participants. Measured variable were, among others, body fat (\%), age, weight, body size, BMI, neck thickness and abdominal girth.  

```{r, echo = FALSE, message=FALSE, warning=FALSE}
library(dplyr)
path <- "../../../3_datasets/"
d.bodyfat <- read.table(paste(path,"bodyfat.clean.txt",sep=""),header=T)
d.bodyfat <- d.bodyfat[,c("bodyfat","age","gewicht","hoehe","bmi","neck","abdomen")]
```

```{r echo = TRUE}
glimpse(d.bodyfat)
```

## Scatterplots

\setkeys{Gin}{width=0.85\textwidth}

```{r, fig.width=6, fig.height=6, out.width="6cm", fig.align = 'center'}
pairs(d.bodyfat)
```

## 


```{r echo = FALSE, fig=TRUE, out.width = "6cm", out.height='6cm', fig.align='center', include = TRUE}
par(mfrow=c(2,2),mar=c(4,4,1,1))
plot(bodyfat ~ bmi ,data=d.bodyfat)
plot(bodyfat ~ gewicht ,data=d.bodyfat,ylab="")
plot(bodyfat ~ neck ,data=d.bodyfat)
plot(bodyfat ~ abdomen ,data=d.bodyfat,ylab="")
```

We are looking for a _model_ that \alert{predicts} body fat as precisely as possible from variables that are easy to measure. 

## Data example 2: Mercury (Hg) in Valais (Switzerland)

__Question:__ Association between Hg concentrations in the soil and in urine of the people living in the respective properties. We use a slightly modified data set here.

```{r echo = FALSE}
#path <- "../../../3_datasets/"
#d.hg <- read.table(paste(path,"Hg_urin.csv",sep=""),header=T, sep=",")
#d.hg <- d.hg[,-c(11)]
#names(d.hg) <- c("Hg_urin", "Hg_soil", "veg_garden", "migration", "smoking","amalgam", "age", "fish", "last_time_fish", "mother")
#d.hg$mother <- as.factor(d.hg$mother)
#set.seed(234)
#d.hg$Hg_soil <- d.hg$Hg_soil[sample(1:nrow(d.hg))]
```

```{r echo = TRUE}
#glimpse(d.hg)
```

## 

A first visual inspection is not very informative. There is not much that is visible by eye:

```{r,echo=FALSE, out.width='3.5cm', out.height='3.5cm', fig.pos='center', message=FALSE, warning=FALSE}
library(ggplot2)
#ggplot(d.hg,aes(x=Hg_soil,y=Hg_urin)) + geom_point() + 
#  theme_bw() +
#  xlab("Hg soil") +
#  ylab("Hg urine")
```

## 

Which other factors might be responsible for high Hg concentrations in urine?

```{r echo=FALSE, out.width='5.5cm', out.width='2cm', fig.pos='center', message=FALSE, warning=FALSE}
require(cowplot)
#theme_set(theme_cowplot(font_size=10)) # reduce default font size

#plot.1 <- ggplot(d.hg,aes(x=amalgam,y=Hg_urin)) + geom_point(size=0.5) + 
#  theme_bw() +
#  xlab("Amalgam") +
#  ylab("Hg urine")

#plot.2 <- ggplot(d.hg,aes(x=age,y=Hg_urin)) + geom_point(size=0.5) + 
#  theme_bw() +
#  xlab("Age") +
#  ylab("Hg urine")

#plot.3 <- ggplot(d.hg,aes(x=fish,y=Hg_urin)) + geom_point(size=0.5) + 
#  theme_bw() +
#  xlab("Fish consumption") +
#  ylab("Hg urine")

#plot_grid(plot.1, plot.2, plot.3, ncol=3)
```

From these plots it is hard to tell which factors exactly influence the Hg pollution in humans.

## 

It is always useful to look at the distribution of the variables in the model. Let us plot the histogram of Hg concentrations:  

```{r out.height='2cm', out.width='4cm', echo=FALSE, fig.align='center'}
#plot.1 <- ggplot(d.hg,aes(x=Hg_soil)) + geom_histogram() + xlab("Hg soil")
#plot.2 <- ggplot(d.hg,aes(x=Hg_urin)) + geom_histogram() + xlab("Hg urine")
#plot_grid(plot.1, plot.2,ncol=2)
```

All Hg values seem to "stick" at 0.

## 

The scatterplot does also look much more reasonable with log-transformed values:

```{r out.height='4.5cm', out.width='4.5cm', echo=FALSE, fig.align='center'}
#plot(log(Hg_urin) ~ log(Hg_soil), data=d.hg, xlab="log(Hg soil)", 
#     ylab = "log(Hg Urin)",pch=21,bg=as.numeric(mother)+2,xlim=c(-4.5,4.5))
#legend("topright",legend=c("Children","Mothers"),col=c(3,4),pch=21,pt.bg=c(3,4))
```

Remember: The idea to log-transform the variables was mainly obvious thanks to \alert{visual inspection}! 

## Data example 3: Diet and blood glucose level

<!--{\scriptsize\citep[p. 190]{elpelt.hartung1987}}-->

24 persons were split into 4 groups. Each group followed another diet (DIAET). The blood glucose concentrations were measured at the beginning and at the end (after 2 weeks). The difference of these values was stored (BLUTZUCK).  

\colorbox{lightgray}{\begin{minipage}{14cm}
{\bf Question:} Are there differences among the groups with respect to changes in blood glucose concentrations?
\end{minipage}}

Let's look at the raw data (points and boxplots):

```{r echo = FALSE}
path <- "../../../3_datasets/"
d.blz <- read.table(paste(path,"blutzucker.dat",sep=""),header=T)
```

```{r out.height='4cm', out.width='4cm',echo=FALSE, fig.align='center'}
ggplot(d.blz,aes(x=factor(DIAET),y=BLUTZUCK)) + geom_boxplot() + geom_point(size=3,colour='grey') + theme_bw() + xlab("Diet") + ylab("Blood glucose")
```

## Data example 4: Blood-screening

<!--{\scriptsize\citep[From ][Chapter 7.1]{hothorn.everitt2014}}-->

Is a high ESR (erythrocyte sedimentation rate) an indicator for certain diseases (rheumatic disease, chronic inflammations)?  

__Specifically:__ Is there an association between the concentrations of the plasma proteins Fibrinogen and Globulin and ESR level ESR$<20mm/hr$?

The `plasma` data come with the package `HSAUR3` <!--\citep{hothorn.everitt2014}.-->

```{r echo = FALSE, message=FALSE, warning=FALSE}
library("HSAUR3")
data("plasma",package="HSAUR3")
plasma$status <- factor(ifelse(plasma$ESR=="ESR < 20","healthy","sick"))
plasma[c(1,5,9,10,15,29),]
```

## 

The distinction ESR$<20mm/hr$ (healthy) vs.\ ESR$\geq 20mm/hr$ (sick) leads to a \alert{binary} response variable.  

\begin{center}$\rightarrow$ \emph{conditional density plot}\end{center}

```{r out.width='7cm', out.height='4cm', fig.align='center'}
par(mfrow=c(1,2))
cdplot(status ~ fibrinogen,plasma)
cdplot(status ~ globulin,plasma)
```

## What is a model?

A model is an approximation of the reality. __Understanding how the real world works__ is usually only possible thanks to simplifying assumptions.

$\rightarrow$ This is exactly __the purpose of statistical data analysis.__

In 2014, David Hand wrote:

_In general, when building statistical models, we must
not forget that the aim is to understand something about
the real world. Or predict, choose an action, make
a decision, summarize evidence, and so on, but always
about the real world, not an abstract mathematical
world: our models are not the reality -- a point well
made by George Box in his often-cited remark that_
\alert{"all
models are wrong, but some are useful."}
<!--\citep{box1979}.}_-->

## Steps in a modelling process ("work flow")

\begin{center}
\includegraphics[width=11cm]{pictures/workflow.png}
\end{center}

## The scopes of statistical data analysis

a. \alert{Prediction (extrapolation), interpolation}. Example body fat: use substitute measurements to predict body fat of a person.

b. \alert{Explanation;  determination of important variables}. Example physical activity of children: The study aims to find factors that (positively or negatively) influence the movement behavior of children.

c. \alert{Estimation of parameters and quantify the uncertainty.} Example: Effect size of a novel drug.

d. Optimization.

e. Calibration.

In this course we are concerned with a-c.

## Goals of the course (part 2)

By the end of the course you will be able

* to \textcolor{blue}{analyze} all data examples introduced here using R (and of course many more),
* to \textcolor{blue}{report and interpret} the results,
* to \textcolor{blue}{draw conclusions} from them,
* to give \textcolor{blue}{graphical descriptions} of the data and the results,
* to \textcolor{blue}{be critical} about what you see.

## Graphical representation of data

You should remember the following options for graphical data descriptions. Several of them appeared already in previous examples.

\begin{tabular}{ll}
Representation & Useful for \\
\hline
 Scatterplots & Pairwise dependency of continuous \\
    &  variables. \\[1mm]
 Histograms & Distribution of continuous variables.\\[1mm]
 Boxplots  & Distribution of continuous variables, ev. \\
    & conditionally on categories.\\[1mm]
 Conditional density plots &  Dependency of a binary variable from\\
    &  a continuous variable.\\[1mm]
%Barplots & As  (see chapter 10.1 in the Mat183 script)boxplots. \\[1mm]
Coplots & Dependencies among multiple variables. \\
\hline
\end{tabular}

## Coplots

Ideal to graphically display dependencies when more than two variables are involved. Very useful for categorical variables. Example: Mercury in Valais.

```{r out.width='6cm', out.height='5cm'}
#coplot(log(Hg_urin) ~  age | mother * migration,
# d.hg, panel = panel.smooth)
```

##

There are many "fancy" ways to graphically display data (__nice-to-know__):

* 3D-plots
* Spatial representations (using geodata)
* Interactive graphs and animations

Many R packages are available for various purposes. Interactive apps can, for example, be generated with Shiny. Check out the shiny gallery:

\url{http://shiny.rstudio.com/gallery/}

## Next week: Simple linear regression

It will be partially a repetition of what you heard in Mat183, chapter 10.2.

## 

References

Box, G.E.P. (1979). Robustness in the strategy of scientific model building. In R.L. Launer and G.N. Wilkinson (Eds.), _In Robustness in Statistics_, pp. 201-236. New York: Academic Press.

Elpelt, B. and J. Hartung (1987). _Grundkurs Statistik, Lehr- und Übungsbuchder angewandten Statistik.

Hothorn, T. and B.S. Everitt (2014). _A Handbook of Statistical Analyses Using R_ (3 ed.). Boca Raton: Chapman & Hall/CRC Press.
<!--\bibliographystyle{Chicago}-->
<!--\bibliography{refs}-->